Active Learning for SAT Solver Benchmarking

### 2022-10-10 paper/paper.tex
Revise Section 3.2 (Solver Model)
E.g., make it a bit shorter and improve argumentation for discretization.

Our framework to score a new solver~$\hat{a}$ follows a simple three-step procedure, outlined in Algorithm~\ref{algALBenchmark}:
While the stopping condition is not met (line~1), we select \emph{one} problem instance from the pool of instances with the help of a prediction model~$\mathcal{M}$ (line~2) and evaluate the new solver~$\hat{a}$ on this instance (line~3).
We then use the acquired result to update the prediction model~$\mathcal{M}$ (line~4).
When the stopping criterion is met, we return the predicted score of the new solver~$\hat{a}$ (line~5).

@@

Let benchmark instances $\mathcal{I}$, SAT solvers $\mathcal{A}$, and runtimes $r : \mathcal{A} \times \mathcal{I} \rightarrow \left[0, \tau\right]$ be given. 
We denote the new solver to be ranked by $\hat a \not\in A$ and define $\hat A := A \cup \hat a$ to denote all solvers including the new solver. 
At the core of the model $\mathcal{M}$ is a prediction function $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \mathbb{R}$ that powers the decisions within the previously mentioned three steps.

To be applicable to $\hat{a}$, we extend \emph{score} by replacing $\gamma(a, e)$ by the model's predictions $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$.

@@

We transform the runtimes into discrete runtime classes which are determined on a per instance basis.
For each instance $e \in \mathcal{I}$, we use a single-link hierarchial clustering algorithm to assign the runtimes in $\bigl\{ r(a, e) \mid a \in A \bigr\}$ to one of the $k$ clusters $C_1, \dots, C_k$ such that the fastest runtimes for instance $e$ are in cluster $C_1$ and the slowest are in cluster $C_k$.
For instances with many timeouts, we have to manually fix the clustering results such that all timeouts $\tau$ are in cluster $C_k$.
The runtime transformation function $\gamma_k : {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ is then specified as follows.
$$\gamma_k(a, e) = j ~\Leftrightarrow~ r(a, e) \in C_j$$

Given an instance $e \in \mathcal{I}$, a solver $a \in A$ belongs to the $\gamma_k(a, e)$-fastest solvers on instance $e$. 
In our preliminary experiments, we achieved higher accuracy with prediction of such discrete runtime labels than with prediction of raw runtimes. 
The mean squared error of raw runtime prediction was within the same magnitude as the values to be predicted. 
%MI: how much more accurate are we with discrete runtime labels
Research in portfolio solvers has also shown that directly predicting runtimes without any transformation does not work well in practice~\cite{NgokoCT19,CollauttiMMO13}.

To determine solver ranks, we use the transformed runtimes $\gamma_k(a, e)$ in the adapted scoring function $s_k : A \rightarrow [1, 2 \cdot k]$ as follows.
\begin{align}
  s_k(a) := \frac{1}{|\mathcal{I}|} \sum_{e \in \mathcal{I}} \gamma'_k(a, e)
  &&
  \gamma'_k(a, e) := \begin{cases}
  	2 \cdot \gamma_k(a, e) 	& \text{if } \gamma_k(a, e) = k\\
	\gamma_k(a, e)	& \text{otherwise}
  \end{cases}
  \label{eq:rankingeq}
\end{align}

The scoring function $s_k$ induces a ranking among solvers. 
In preliminary experiments, we compared the ranking induced by $s_3$ to the ranking induced by the PAR-2 score which is commonly used in SAT competitions~\cite{FroleyksHIJS21}. 
The Spearman correlation of $s_3$ and PAR-2 ranking is about \SI{0.988} which is very close to the optimal value of 1~\cite{de2016comparing}.

That discretized runtimes are sufficient for discriminating solvers can also be shown by comparing the fraction of solver pairs whose runtimes are significantly different. 
In preliminary experiments, we used a Wilcoxon signed rank test with $\alpha = 0.05$ to compare solver pairs by their runtime vectors.
Within the SAT~Competition~2022 Anniversary track, about \SI{89.95}{\%} of solver pairs are significantly different according to that measure.
The strongly discretized runtimes $\gamma_2$ are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs, while $\gamma_3$ is sufficient to distinguish about \SI{87.83}{\%} of solver pairs.

% MI: Enough justification, commented out the following
%For the SAT Competition~2022 Anniversary track instances, this ranking approach is capable of correctly deciding for almost all pairs of solvers which one is faster (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$).
%Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.

This shows that monotonicity and score distances between ranks align quite well between both ranking methods.
The correlation between $s_k$-induced and PAR-2 induced ranks, the discriminatory power of $\gamma_k$, as well as that it enables us to use classification, show the clear advantage of runtime discretization.

@@

Selecting an instance based on the model is the core functionality of our method (cf. Algorithm~\ref{algALBenchmark}, Line 2). 
In this section, we introduce our sampling strategies that make use of the model's label-prediction function $f$ which are inspired by existing work within the realms of active learning~\cite{settles2009active}.
We implement a model-uncertainty based approach and a model-based information-gain sampling strategy.

These methods require the model's predictions to also include probabilities for the $k$ possible runtime labels, i.e., 
\begin{equation*}
  f' : \mathcal{\hat A} \times \mathcal{I} \rightarrow \left[0, 1\right]^k
\end{equation*}

@@

The model-uncertainty-based approach simply selects the instance that is closest to the model's decision boundary and, of course, not already selected, i.e.,
\begin{equation*}
  \underset{e \in \mathcal{I} \setminus \tilde{\mathcal{I}}}{\arg\min} \left\lvert \frac{1}{k} - \max_{n \in \left\lbrace 1, \dots, k \right\rbrace} f'\!\left(\hat{a}, e\right)_{n} \right\rvert
\end{equation*}

@@

The model-based information-gain sampling strategy selects the instance with the highest expected entropy reduction.
To be more specific, we select the instance $e$ that maximizes
\begin{equation*}
  \operatorname{IG}(e) := \operatorname{H}(e) - \sum_{n = 1}^{k} f'(\hat{a}, e)_{n} \operatorname{H}(e, n)
\end{equation*}
whereby $\operatorname{H}\!\left(i\right)$ denotes the entropy of the runtime labels $\gamma\!\left(a, i\right)$ for all $a \in \mathcal{A}$ and $\operatorname{H}\!\left(i, n\right)$ denotes the entropy of the runtime labels including the runtime label of $\hat{a}$ which is $n$.
The term $\operatorname{H}\!\left(i, n\right)$ is computed for every possible runtime label $n$.

@@

\subsubsection{Wilcoxon Stopping Criterion}
The Wilcoxon stopping criterion stops the iterative active learning algorithm when we are on average certain enough that the predicted runtime labels are sufficiently different in comparison to the other solvers.

To be more specific, we stop when
\begin{equation}
  W_{\hat{a}} := \frac{1}{\left\lvert \mathcal{A} \right\rvert} \sum_{a \in \mathcal{A}} \operatorname{wilcoxon}_{\alpha}\!\left(\gamma\!\left(a, \mathcal{I}\right), f\!\left(\hat{a}, \mathcal{I}\right) \right)
\end{equation}
drops below a certain threshold.
Thereby, \emph{wilcoxon} denotes the $p$-value of a Wilcoxon signed-rank test of the runtime-label distributions and $\gamma\!\left(a, \mathcal{I}\right)$ and $f\!\left(\hat{a}, \mathcal{I}\right)$ are the natural set extensions on $\gamma$ and $f$ respectively.
To improve the stability of this criterion, we use an exponential moving average to smooth out outliers.

%whereby $m$ denotes the count of iterations within Algorithm~\ref{algALBenchmark} and $W_{\exp}^{\left(0\right)} := 1$.
%\begin{equation}
%  W_{\exp}^{\left(m\right)} := \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(m - 1\right)}
%\end{equation}

@@

The model-convergence stopping criterion is less sophisticated in comparison.
It simply stops the active-learning process if within the last $l$ iterations the ranking that is induced by the model's predictions (Equation~\ref{eq:rankingeq}) remains unchanged.
Note that the concrete values of $s_k(a)$ might still change.
We are solemnly interested in the induced ranking in this case.

\section{Experimental Design}

@@

\begin{itemize}
  \item Observed PAR-2 ranking of already sampled instances
  \item Predicted ranking induced by runtime-label predictions
  \begin{itemize}
    \item
    History size: consider the latest 1, 10, 20, 30, or 40 predictions within a voting approach for stability.
    Given a particular ranking decision, the latest $x$ predictions vote on the winner.
    %JB: confuses me a bit as it sounds we rank first and then vote on a winner; isn't it the other way around?
    \item
    Fallback threshold: if the difference of scores between two solvers drops below \SI{0.01}, \SI{0.05}, or \SI{0.1}, use the partially observed PAR-2 ranking as a tie-breaker.
    %JB: as clarification: one these two solvers has to be the new solver, right? if yes, would write it explicitly
    As discussed in Section~\ref{sec:main-model}, ranking decisions are all correct if the difference in scores between two solvers exceeds a certain value.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
  \item Fixed subset size of \SI{10}{\%} or \SI{20}{\%} of instances
  \item Ranking convergence criterion
  \begin{itemize}
    \item Minimum amount: sample at least \SI{2}{\%}, \SI{8}{\%}, \SI{10}{\%}, or \SI{12}{\%} of instances before applying the criterion.
    \item Convergence duration: stop if the predicted ranking stays the same for the last \SI{1}{\%} or \SI{2}{\%} of sampled instances.
    %JB: why this fraction and not an absolute number of instances? the definition here might make stopping in early iterations rather likely, as just very few consecutive instances are needed (which might happen by coincidence; e.g., for ~5k instances, 2% minimum amount-> ~100 instances, 1% convergence duration -> ~1 instance), while the criterion becomes harder in later iterations
    %TF: for that reason, it is combined with the "minimum amount". Using fractions makes the approach more useful when adapting it to benchmarks of different sizes.
    %JB: yes, but (1) even combination with "minimum amount" can make stopping very easy (see my example) and (2) fraction for convergence could also be relative to absolute number of instances (which is constant over AL run) rather than fraction of currently sampled instances (which increases over AL run)
  \end{itemize}

@@

As a runtime-label prediction model, we only make use of one method that has proven to be successful in preliminary experiments (cf.~Section~\ref{sec:evalprel}).
Our model of choice is an ensemble, stacking a quadratic-discriminant analysis\footnote{With a regularization parameter of zero and a singular value threshold of zero} onto a random forest model\footnote{With \emph{entropy} splitting criterion and \emph{balanced} class weights} using a decision tree\footnote{With \emph{Gini} impurity splitting criterion; choosing the best split for each branching decision and maximum tree depth of 5 levels} to weight its members.
If not stated explicitly, all other model parameters default to the presets of \emph{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/index.html}; Version 1.0.2}~\cite{scikit-learn}.

@@

\section{Evaluation}
\label{sec:eval}
Now that we are all set, we want to, first, discuss preliminary results including the choice of the runtime-discretization method and the choice of the machine-learning predictor.
Thereafter, we look at end-to-end results and instance-family importance.
%JB: instance families not really motivated till now, only appear in abstract and in text regarding stratification

\subsection{Preliminary Results}
\label{sec:evalprel}
%JB: if we dedicate a whole section to these results, we should probably rename this section (not "preliminary" in the typical sense); for TACAS, I guess this can be cut a bit
An exhaustive grid search of all approaches outlined in Section~\ref{sec:main} is infeasible.
We are, therefore, narrowing down the search by filtering based on preliminary results.
Since all relevant decisions, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}, revolve around the choice of prediction model, we are first looking at the \emph{Matthews Correlation Coefficient} (MCC) scores~\cite{gorodkin2004comparing,matthews1975comparison} that a particular prediction model achieves.
MCC scores are great in dealing with class imbalances in contrast to conventional metrics like accuracy.

\begin{table}
  \centering
  \caption{
  	Time-out prediction Matthews Correlation Coefficient performances.
  	Results are obtained using cross-validation.
  	We provide \SI{64}{\%} of ground-truth for training, use \SI{16}{\%} as a validation set, and \SI{20}{\%} as a test set (5-fold cross-validation).
  }
  \label{tab:mcc}
  \begin{tabular}{lc}
    Prediction Model & Avg. MCC (Std.) \\
    \hline
    Stacking ensemble (QDA, RF)           & 0.9527 (0.0292) \\
    Quadratic-discriminant analysis (QDA) & 0.9290 (0.0339) \\
    Random forest (RF)                    & 0.8530 (0.0479) \\
    AdaBoost with decision trees          & 0.8384 (0.0444) \\
    AdaBoost with logistic regression     & 0.8158 (0.0985) \\
    Decision tree                         & 0.8059 (0.0707) \\
    Logistic regression                   & 0.8052 (0.1018) \\
    kNN classifier                        & 0.7885 (0.1521) \\
    MLP classifier                        & 0.7760 (0.1408) \\
    Support-vector classifier             & 0.7757 (0.2149) \\
    Naive Bayes                           & 0.7306 (0.1394) \\
    \hline
  \end{tabular}
\end{table}

@@

To obtain the results in Table~\ref{tab:mcc}, we perform cross-validation among all solvers: we assume that all other solver runtimes are known and provide \SI{64}{\%} of ground-truth runtime labels of the target solver for training, use \SI{16}{\%} as a validation set for hyper-parameter tuning, and \SI{20}{\%} as a test set (5-fold cross-validation).
This is repeated for each solver once.
Note that we perform cross-validation on two levels, i.e., among the solvers and the target solver runtimes.
We report the average performances on the test sets.
Since the differences in the MCC-performance distributions of the best and second-best approaches are statistically significant regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$, we only use the best approach for all further experiments as a prediction model.

As a runtime discretization technique, we use hierarchical clustering with $k = 3$ using a log-single-link criterion for all further experiments.
Each non-time-out runtime starts in a separate interval.
We then gradually merge intervals whose single-link logarithmic distance is the smallest until the desired number of partitions is reached.
As already mentioned in Section~\ref{sec:main-model}, this approach has good properties regarding the predictiveness of the PAR-2 score ranking and its discriminatory power.
Also, preliminary experiments showed that our stacking ensemble performed particularly well for this kind of discretization method.
Other clustering approaches that we have tried include hierarchical clustering with mean-, median- and complete-link criteria, as well as $k$-means and spectral clustering.
Predicting runtime labels in a similar experiment as above produced an MCC score of on average \SI{88.82}{\%} ($\sigma = \SI{4.75}{\%}$; in comparison to \SI{95.27}{\%} for predicting time-outs).
%JB: "similar experiment as above" - reference unclear; what do you want to say?

\subsection{Experimental Results}
\subsubsection{Tuning our algorithm.}
Our main end-to-end experiments follow the evaluation framework that has been introduced in Section~\ref{sec:evalframe}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches listed in Section~\ref{sec:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the smaller hyper-parameter-tuning dataset.
The plotted lines represent the Pareto front for each possible choice of ranking approaches (Fig.~\ref{fig:annitraincolorranking}), selection approaches (Fig.~\ref{fig:annitraincolorselection}), and stopping criteria (Fig.~\ref{fig:annitraincolorstopping}).
%JB: as far as I understand it, the lines only compare some top-level hyperparameters, but for each of these hyperparemeters, take the best (= Pareto front) out of some more fine-grained settings (the hyperparameters of the hyperparameters, so to speak), so it's no really "each possible choice" in the sense of actual, individual configurations, but rather a combination of configurations
Colors indicate the different hyper-parameter instantiations.
To be more specific, they represent the Pareto front of all configurations with the respective instantiations.

@@

\subsubsection{Instance-family importance.}
For all configurations of our approach that sample on average less than \SI{10}{\%} of all 5355 instances, we look at the particular instances and families that are selected by our active-learning approach.
Thereby, the average benchmark set contains about \SI{192.34}{} instances.
Table~\ref{tab:familiesa} lists the ten most frequent families that are chosen.
For example, \emph{planning} instances account for on average \SI{19.14}{} of the \SI{192.34}{} benchmark instances.
In contrast to that, Table~\ref{tab:familiesb} shows the families that are sampled the most often relative to their occurrences in the underlying dataset.
There are, for example, only nine \emph{circuit-multiplier} instances in the complete dataset.
However, we sample on average \SI{1.59}{} of them.
%JB: there should be some (qualitative) interpretation (based on our results, is the sample interesting/special family-wise or not?) rather than just dropping a table into the paper and mainly explaining its structure

>>>>

Our framework to score a new solver~$\hat{a}$ follows a three-step procedure, outlined in Algorithm~\ref{algALBenchmark}:
While the stopping condition is not met (line~1), we select \emph{one} SAT instance from the pool of instances with the help of a prediction model~$\mathcal{M}$ (line~2) and evaluate the new solver~$\hat{a}$ on this instance (line~3).
We then use the acquired result to update the prediction model~$\mathcal{M}$ (line~4).
When the stopping criterion is met, we return the predicted score of the new solver~$\hat{a}$ (line~5).

@@

Let benchmark instances $\mathcal{I}$, SAT solvers $\mathcal{A}$, and runtimes $r : \mathcal{A} \times \mathcal{I} \rightarrow \left[0, \tau\right]$ be given. 
We denote the new solver to be ranked by $\hat a \not\in A$ and define $\hat A := A \cup \lbrace \hat a \rbrace$ to denote all solvers including the new solver. 
Model $\mathcal{M}$ provides a prediction function $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \mathbb{R}$ that powers the decisions within our framework.

To be applicable to a new solver $\hat{a}$, whose runtimes are unknown, we extend \emph{score} by replacing $\gamma(a, e)$ with the model's predictions $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$.
%JB: "score" and "gamma" aren't defined yet, so there is nothing to extend; score isn't even defined here, only "s", while later sections still use "score"

@@

We transform the runtimes into discrete runtime labels on a per-instance basis.
For each instance $e \in \mathcal{I}$, we use a single-link hierarchical clustering algorithm to assign the runtimes in $\bigl\{ r(a, e) \mid a \in A \bigr\}$ to one of the $k$ clusters $C_1, \dots, C_k$ such that the fastest runtimes for instance $e$ are in cluster $C_1$ and the slowest are in cluster $C_k$.
For instances with many timeouts, we have to manually fix the clustering results such that all timeouts $\tau$ are in cluster $C_k$.
%TF: maybe a bit unclear. We do not manually fix anything. There is always one cluster with only time-outs if there are any.
%JB: was wondering about that as well; how do we exactly create the timeout cluster? do we only cluster with k-1, removing timeout instances, or is the timeout cluster an ordinary cluster (which might also contain on-timeout instances)?
The runtime transformation function $\gamma_k : {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ is then specified as follows:
$$\gamma_k(a, e) = j ~\Leftrightarrow~ r(a, e) \in C_j$$

Given an instance $e \in \mathcal{I}$, a solver $a \in A$ belongs to the $\gamma_k(a, e)$-fastest solvers on instance $e$. 
In preliminary experiments, we achieved higher accuracy predicting such discrete runtime labels than predicting raw runtimes. 
The mean squared error of raw runtime regression was within the same magnitude as the values to be predicted. 
%MI: how much more accurate are we with discrete runtime labels
Research on portfolio solvers has also shown that directly predicting runtimes without any transformation does not work well in practice~\cite{CollauttiMMO13,NgokoCT19}.
%JB: maybe we should phrase this more defensively or just leave out this argument (might be dataset-dependend), (1) e.g., SATzilla 2008 does predict runtimes and (2) we could also have treid to predict normalized runtimes like Collautti et al. instead of discretizing (though prediction performance would probably still have been bad in our case)

To determine solver ranks, we use the transformed runtimes $\gamma_k(a, e)$ in the adapted scoring function $s_k : A \rightarrow [1, 2 \cdot k]$ as follows:
%
\begin{align}
  s_k(a) := \frac{1}{|\mathcal{I}|} \sum_{e \in \mathcal{I}} \gamma'_k(a, e)
  &&
  \gamma'_k(a, e) := \begin{cases}
  	2 \cdot \gamma_k(a, e) 	& \text{if } \gamma_k(a, e) = k\\
	\gamma_k(a, e)	& \text{otherwise}
  \end{cases}
  \label{eq:rankingeq}
\end{align}
%
I.e., we apply a PAR-2 scoring, which is commonly used in SAT competitions~\cite{FroleyksHIJS21}, on the discrete labels.
The scoring function $s_k$ induces a ranking among solvers.

However, we need to ensure that discretized labels still discriminate solvers and align with the actual PAR-2 ranking.
We analyzed the ranking induced by $s_3$ in preliminary experiments with the SAT Competition~2022 Anniversary Track~\cite{sat2022}.
According to a Wilcoxon-signed-rank test with $\alpha = 0.05$, \SI{87.83}{\%} of solver pairs have significantly different scores after discretization, only a slight drop compared to \SI{89.95}{\%} before discretization.
%While only time-out labels (solver times out or not on a particular instance) are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs.
%\footnote{Based on hierarchical clustering with a log-single-link criterion for $k = 3$.}
Further, our ranking approach is capable of correctly deciding for almost all (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$) solver pairs which solver is faster.
%Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
In particular, the Spearman correlation of $s_3$ and PAR-2 ranking is about \SI{0.988}{}, which is very close to the optimal value of 1~\cite{de2016comparing}.
All these results show that discretized runtimes are suitable for our framework.

@@

Selecting an instance based on the model is the core functionality of our method (cf. Algorithm~\ref{algALBenchmark}, Line 2). 
In this section, we introduce our sampling strategies that make use of the model's label-prediction function $f$ which are inspired by existing work within the realms of active learning~\cite{settles2009active}.
We implement a model-uncertainty based approach and a model-based information-gain sampling strategy.
These methods require the model's predictions to also include probabilities for the $k$ possible runtime labels denoted by $f' : \mathcal{\hat A} \times \mathcal{I} \rightarrow \left[0, 1\right]^k$.

@@

The model-uncertainty-based approach simply selects the instance that is closest to the model's decision boundary.

\begin{equation*}
  \underset{e \in \mathcal{I} \setminus \tilde{\mathcal{I}}}{\arg\min} \left\lvert \frac{1}{k} - \max_{n \in \left\lbrace 1, \dots, k \right\rbrace} f'\!\left(\hat{a}, e\right)_{n} \right\rvert
\end{equation*}

@@

The model-based information-gain sampling strategy selects the instance with the highest expected entropy reduction.
To be more specific, we select the instance $e$ that maximizes $IG(e)$ which is specified as follows.

\begin{equation*}
  \operatorname{IG}(e) := \operatorname{H}(e) - \sum_{n = 1}^{k} f'(\hat{a}, e)_{n} \operatorname{H}(e, n)
\end{equation*}

Here $\operatorname{H}(e)$ denotes the entropy of the runtime labels $\gamma(a, e)$ for all $a \in \mathcal{A}$ and $\operatorname{H}(e, n)$ denotes the entropy of the runtime labels including the runtime label of $\hat{a}$ which is $n$.
The term $\operatorname{H}(e, n)$ is computed for every possible runtime label $n \in \{1, \dots, k\}$.

@@

\subsubsection{Wilcoxon Stopping Criterion}

The Wilcoxon stopping criterion stops the iterative active learning algorithm when we are on average certain enough that the predicted runtime labels are sufficiently different in comparison to the other solvers.

To be more specific, we use the average $p$-value $W_{\hat{a}}$ of a Wilcoxon signed-rank test $w(S,P)$ of the two runtime label distributions $S=\{ \gamma(a, e) \mid e \in \mathcal{I} \}$ and $P=\{ f(\hat a, e) \mid e \in \mathcal{I} \}$.

\begin{equation*}
  W_{\hat{a}} := \frac{1}{\lvert \mathcal{A} \rvert} \sum_{a \in \mathcal{A}} \operatorname{w}(S, P)
\end{equation*}

To improve the stability of this criterion, we use an exponential moving average to smooth out outliers as follows and stop as soon as $W^{(i)}_{\hat{a}}$ drops below a certain threshold.

\begin{align*}
	W_{\exp}^{\left(0\right)} &:= 1\\
	W_{\exp}^{\left(i\right)} &:= \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(i - 1\right)}
\end{align*}

@@

The model-convergence stopping criterion is less sophisticated in comparison.
It simply stops the active-learning process if within the last $l$ iterations the ranking that is induced by the model's predictions (Equation~\ref{eq:rankingeq}) remains unchanged.
Note that the concrete values of $s_k(\hat a)$ might still change.
We are solemnly interested in the induced ranking in this case.

\section{Experimental Design}
\label{sec:exdesign}

@@

\begin{itemize}
  \item Observed PAR-2 ranking of already sampled instances
  \item Predicted ranking induced by runtime-label predictions
  \begin{itemize}
    \item
    History size: consider the latest 1, 10, 20, 30, or 40 predictions within a voting approach for stability.
    The latest $x$ predictions vote on the winner.
    %JB: confuses me a bit as it sounds like we rank first and then vote on a winner; isn't it the other way around?
    %TF: Removed confusing sentence
    \item
    Fallback threshold: if the difference of scores between two solvers drops below \SI{0.01}, \SI{0.05}, or \SI{0.1}, use the partially observed PAR-2 ranking as a tie-breaker.
    %JB: as clarification: one these two solvers has to be the new solver, right? if yes, would write it explicitly
    %TF: Not necessarily
    As discussed in Section~\ref{sec:main-model}, ranking decisions are all correct if the difference in scores between two solvers exceeds a certain value.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
  \item Fixed subset size of \SI{10}{\%} or \SI{20}{\%} of instances
  \item Ranking convergence criterion
  \begin{itemize}
    \item Minimum amount: sample at least \SI{2}{\%}, \SI{8}{\%}, \SI{10}{\%}, or \SI{12}{\%} of instances before applying the criterion.
    \item Convergence duration: stop if the predicted ranking stays the same for the last \SI{1}{\%} or \SI{2}{\%} of sampled instances.
    %JB: why this fraction and not an absolute number of instances? the definition here might make stopping in early iterations rather likely, as just very few consecutive instances are needed (which might happen by coincidence; e.g., for ~5k instances, 2% minimum amount-> ~100 instances, 1% convergence duration -> ~1 instance), while the criterion becomes harder in later iterations
    %TF: for that reason, it is combined with the "minimum amount". Using fractions makes the approach more useful when adapting it to benchmarks of different sizes.
    %JB: yes, but (1) even combination with "minimum amount" can make stopping very easy (see my example) and (2) fraction for convergence could also be relative to absolute number of instances (which is constant over AL run) rather than fraction of currently sampled instances (which increases over AL run)
    %TF: I'm sure that both would make sense. Isn't 1% convergence duration of 5k about ~50 instances?
  \end{itemize}

@@

As runtime-label prediction model, we only make use of one fixed model since an exhaustive grid search would be infeasible (cf.~Section~\ref{sec:evalprel}).
Our model of choice is an ensemble, stacking a quadratic-discriminant analysis\footnote{With a regularization parameter of zero and a singular value threshold of zero} onto a random forest model\footnote{With \emph{entropy} splitting criterion and \emph{balanced} class weights} using a decision tree\footnote{With \emph{Gini} impurity splitting criterion; choosing the best split for each branching decision and maximum tree depth of 5 levels} to weight its members.
If not stated explicitly, all other model parameters default to the presets of \emph{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/index.html}; Version 1.0.2}~\cite{scikit-learn}.

@@

\section{Evaluation}
\label{sec:eval}

Before evaluating our approach end-to-end as described in Section~\ref{sec:exdesign}, we discuss the choice of the runtime-discretization method and the machine-learning predictor.
Thereafter, we look at the results of our active-learning approach.
Selection decisions of the aforementioned approach also reveal the importance of different problem-instance families to our model.
Instance families comprise instances that are derived from the same application domain, e.g., planning, cryptography, etc., and are a useful tool for both analyzing solver performance and portfolio creation.
%JB: instance families not really motivated till now, only appear in abstract and in text regarding stratification
%TF: Added a sentence for clarification. We could introduce more details in section 4.3.

\subsection{Runtime Prediction}
\label{sec:evalprel}

%JB: if we dedicate a whole section to these results, we should probably rename this section (not "preliminary" in the typical sense); for TACAS, I guess this can be cut a bit
%TF: cut the section down a bit, renamed to "runtime prediction"
An exhaustive grid search of all active-learning approaches outlined in Section~\ref{sec:main} and different runtime-prediction models is infeasible.
Therefore, we use the runtime-prediction model that produced the best results for all further experiments.
We compare runtime predictors by looking at the \emph{Matthews Correlation Coefficient} (MCC) scores~\cite{gorodkin2004comparing,matthews1975comparison} of the classification task.
MCC scores are great in dealing with class imbalances in contrast to conventional metrics like accuracy.

% \begin{table}
%   \centering
%   \caption{
%   	Time-out prediction Matthews Correlation Coefficient performances.
%   	Results are obtained using cross-validation.
%   	We provide \SI{64}{\%} of ground-truth for training, use \SI{16}{\%} as a validation set, and \SI{20}{\%} as a test set (5-fold cross-validation).
%   }
%   \label{tab:mcc}
%   \begin{tabular}{lc}
%     Prediction Model & Avg. MCC (Std.) \\
%     \hline
%     Stacking ensemble (QDA, RF)           & 0.9527 (0.0292) \\
%     Quadratic-discriminant analysis (QDA) & 0.9290 (0.0339) \\
%     Random forest (RF)                    & 0.8530 (0.0479) \\
%     AdaBoost with decision trees          & 0.8384 (0.0444) \\
%     AdaBoost with logistic regression     & 0.8158 (0.0985) \\
%     Decision tree                         & 0.8059 (0.0707) \\
%     Logistic regression                   & 0.8052 (0.1018) \\
%     kNN classifier                        & 0.7885 (0.1521) \\
%     MLP classifier                        & 0.7760 (0.1408) \\
%     Support-vector classifier             & 0.7757 (0.2149) \\
%     Naive Bayes                           & 0.7306 (0.1394) \\
%     \hline
%   \end{tabular}
% \end{table}

@@

%To obtain the results in Table~\ref{tab:mcc},
We perform cross-validation among all solvers: we assume that all other solver runtimes are known and provide \SI{64}{\%} of ground-truth runtime labels of the target solver for training, use \SI{16}{\%} as a validation set for hyper-parameter tuning of the runtime predictor, and \SI{20}{\%} as a test set (5-fold cross-validation).
This is repeated for each solver once.
Note that we perform cross-validation on two levels, i.e., among the solvers and the target solver runtimes.
We report the average performances on the test sets.

ML classification models that we have tried include random forests, decision trees, quadratic-discriminant analysis, AdaBoost, logistic regression, $k$-nearest neighbor, multi-layer perceptrons, support vector machines, and naive Bayes.
A stacking ensemble consisting of quadratic-discriminant analysis and random forests performed best.
It produces an MCC score of \SI{0.9527}{} for time-out prediction and \SI{0.8882}{} for runtime-label prediction.
The results are significantly better than any other model that we have tried regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$.

%Since the differences in the MCC-performance distributions of the best and second-best approaches are statistically significant regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$, we only use the best approach for all further experiments as a prediction model.

For runtime discretization, we use hierarchical clustering with $k = 3$ using a log-single-link criterion for all further experiments.
Each non-time-out runtime starts in a separate interval.
We then gradually merge intervals whose single-link logarithmic distance is the smallest until the desired number of partitions is reached.
As already mentioned in Section~\ref{sec:main-model}, this approach has good properties regarding the predictiveness of the PAR-2 score ranking and its discriminatory power.
Also, experiments showed that our stacking ensemble performed particularly well for this kind of discretization method.
Other clustering approaches that we have tried include hierarchical clustering with mean-, median- and complete-link criteria, as well as $k$-means and spectral clustering.
%Predicting runtime labels in a similar experiment as above produced an MCC score of on average \SI{88.82}{\%} ($\sigma = \SI{4.75}{\%}$; in comparison to \SI{95.27}{\%} for predicting time-outs).
%JB: "similar experiment as above" - reference unclear; what do you want to say?
%TF: Removed the sentence

\subsection{Experimental Results}
\subsubsection{Tuning our algorithm.}

Our main end-to-end experiments follow the evaluation framework that has been introduced in Section~\ref{sec:evalframe}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches listed in Section~\ref{sec:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the smaller hyper-parameter-tuning dataset.
The plotted lines represent the best performing configurations per ranking approach (Fig.~\ref{fig:annitraincolorranking}), per selection approach (Fig.~\ref{fig:annitraincolorselection}), and per stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).
%JB: as far as I understand it, the lines only compare some top-level hyperparameters, but for each of these hyperparemeters, take the best (= Pareto front) out of some more fine-grained settings (the hyperparameters of the hyperparameters, so to speak), so it's no really "each possible choice" in the sense of actual, individual configurations, but rather a combination of configurations
Colors indicate the different hyper-parameter instantiations.
To be more specific, they represent the Pareto front of all configurations with the respective instantiations.

@@

\subsection{Instance-Family Importance}

For all configurations of our approach that sample on average less than \SI{10}{\%} of all 5355 instances, we look at the particular instances and families that are selected by our active-learning approach.
Thereby, the average benchmark set contains about \SI{192.34}{} instances.
Table~\ref{tab:familiesa} lists the ten most frequent families that are chosen.
For example, \emph{planning} instances account for on average \SI{19.14}{} of the \SI{192.34}{} benchmark instances.
In contrast to that, Table~\ref{tab:familiesb} shows the families that are sampled the most often relative to their occurrences in the underlying dataset.
There are, for example, only nine \emph{circuit-multiplier} instances in the complete dataset.
However, we sample on average \SI{1.59}{} of them.
%JB: there should be some (qualitative) interpretation (based on our results, is the sample interesting/special family-wise or not?) rather than just dropping a table into the paper and mainly explaining its structure
